Modelling Prehistoric Iconographic Compositions. The \proglang{R} package \pkg{iconr}

### 2021-01-19 article_rvTH18.Rnw
up

%% This modifies which proportion of the page must be occupied by the figure
%% or table to take the full page (with no additional text).
%% Check if the journal does not have any rule about not changing this.
\renewcommand{\floatpagefraction}{.8}

% \setlist[itemize]{noitemsep, topsep=0pt}
<<preliminaries, echo=FALSE, include=FALSE, cache=FALSE >>=
# options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
library("MASS")
library("extrafont") #;library("ggplot")
library("igraph")
opts_chunk$set(prompt=TRUE, fig.pos = '!htbp', results='asis', highlight=FALSE,
               class.source = "bg-success")
options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
options(SweaveHooks=list(fig=function() par(mar=c(0, 0, 0, 0))))
# options(replace.assign=TRUE, width=90, prompt="R> ")
opts_knit$set(concordance=TRUE, self.contained=FALSE, background = "#FFFFFF")
# knitr::opts_chunk$set(fig.pos = '!htbp')
#loadfonts(device = "win")
loadfonts(device = "pdf")
@

@@

\begin{document}
%\SweaveOpts{concordance=TRUE}
%\SweaveOpts{concordance=TRUE}
% \SweaveOpts{concordance=TRUE}
% \Sweave2knitr("article_rvTH14_1.Rnw")
% \SweaveOpts{concordance=TRUE}

@@

Symbolic practices is a characteristic trait of human societies. Even discussed, such practices seems to start between 233,000 to 800,00 BC \citep{dErrico00}, covering more than 97\% of total human societies time span. Symbolism cover a large range of practices, from ochre deposit in a tomb, to menhir alignements, passing through wall fresco. This latter, what might be called "iconographical practices", probably shows the most complex and interesting testimonies of past societies symbolism. For decades, its study was linked to history of religion because commonly seen as closely linked to cultual practices and believes. Since the \textit{New Archaeology} developpement during the 60's \citep{Clarke14}, symbolic expressions start to be studied with the same formal methods (statistics, seriations, distribution maps, etc.) as any another aspect of social organisation: settlement patterns, tools \emph{chaine op\'eratoire}, susbsitence strategies, etc. \citep{Renfrew91}. But unlike many aspects of the material culture where technological requirements and technical efficiency determine the choice of the raw material and the shape of the object -- a flint blade for cutting, a pottery for containing, a house for living --, the function of an iconographic composition cannot be drawn directly from itself. Furthermore, there is no specific and controled vocabularies to describe hand-made images. There is always a difficulty to decribe an iconographical content with alphanumerical signs. Whether these last decades study of ancient iconography had  undergone significative improvements at the site scale -- with GIS/database statistics, paleoenvironmental restitutions, etc. -- and at the sign scale with the development of archaeological sciences -- radiocarbon dating, use-wear analysis, elemental analysis, etc. --, these improvement do not necessarly help to understand the semantic content of the iconography.\\
Semantics or semiotics can be defined as a system of conventional features -- called signs -- organised also in conventional manners. Until our days, formal methods to study ancient iconography semantics have been mostlty been grounded  -- explicitly or not -- on the prime principle of Saussurian linguistic: the '\emph{linearity of the signifier}' \citep{Saussure89}.
Writing is one of the most rational semiographical system with a clear distinction between signified and signifier and the development of the signified on a linear axis. Even if we do not understand the meaning of the signs, writing can easily be modelled with Graph theory and recurent patterns can be identified. For example, the 3-letters word "\code{art}" can be modelled with three vertices (\code{a}, \code{r}, \code{t}) and two edges (one between \code{a} and \code{r}, the other between \code{r} and \code{t}). In \proglang{R}, these features, concatenated in this order with a \code{paste0()}, is \code{art}, and not \code{rat} (Figure \ref{fig:fart}).

<<a_r_t, echo=FALSE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.width=1.25, fig.height=0.25, fig.cap="\\label{fig:fart}Concatenate of 'a', 'r' and 't' graphical units is 'art'">>=
par(mar=c(0.1,0.1,0.1,0.1) )
g <- graph_from_literal(a-+r, r-+t)
plot(g,
     layout = cbind(seq(-1,1,length.out = gorder(g)), 0),
     vertex.color="white",
     vertex.frame.color="white",
     vertex.size=40,
     vertex.label.cex=1,
     vertex.label.color="black",
     vertex.label.family="Courier",
     edge.arrow.size=0.5,
     xlim=c(-0.94,0.94), 
     ylim=c(-1,-1))
@

@@

<<echo=FALSE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.width=1.4, fig.height=1.4, fig.cap="\\label{fig:fratar}Potential spatial relations between graphical units">>=
par(mar=c(0,0,0,0))
ugs <- data.frame(name=c("a","r","t"))
lks <- data.frame(from=c("a", "r", "t"), to=c("r", "t", "a"), type=c("=","=","="))
g <- igraph::graph_from_data_frame(lks, directed=FALSE, vertices= ugs)
plot(g,
     vertex.color="white",
     vertex.frame.color="white",
     vertex.size=80,
     vertex.label.cex=1,
     vertex.label.color="black",
     vertex.label.family="Courier",
     xlim=c(-0.65,0.65), 
     ylim=c(-0.65,0.65)
)
@

Because of the inherent variability of iconography, and because graphical and spatial proximities between graphical units are generally not quantified, applying the Saussurian model to any structured prehistorical graphical content had led to considerable problems:
\begin{itemize}
\item unexplicit spatial groupings of graphical units -- like graphical units grouped into \emph{figures},  \emph{figures} grouped into \emph{patterns}, \emph{patterns} grouped into \emph{motives}, etc. -- with tedious number of groups
\item consistency, proximities and relationships between these groups are often implicit and not quantified
\item studies develop proper descriptive vocabularies, singular relationships of grouping, idosyncratic methods at site-dependent or period-dependent scales
\end{itemize}

@@

Whether this theoretical process includes image binarization, polygonization and \emph{birel} analysis, a simplier solution will be to create directly centroids on the GUs and to draw the edges manually. Then \emph{x} and \emph{y} coordinates of nodes can be easily retrieved.\\

@@

<<dec.1.4, echo=FALSE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.width=15, fig.asp=0.5630631, fig.cap="\\label{fig:dec.1.4} Cerro Muriano 1 and Zarza de Montanchez decorations.">>=
library(iconr)
par(mar=c(0, 0, 0, 0))
nds.df <- read_nds(site = "Cerro Muriano", 
                   decor = "Cerro Muriano 1",
                   doss = system.file("extdata", 
                                      package = "iconr"))
eds.df <- read_eds(site = "Cerro Muriano",
                   decor = "Cerro Muriano 1",
                   doss = system.file("extdata", 
                                      package = "iconr"))
cm1 <- plot_dec_grph(nodes = nds.df,
                     edges = eds.df,
                     site = "Cerro Muriano",
                     decor = "Cerro Muriano 1",
                     nd.var = "type",
                     lbl.size = 1.7,
                     doss = system.file("extdata", 
                                        package = "iconr"))
cm1.img <- magick::image_read(cm1)
nds.df <- read_nds(site = "Zarza de Montanchez", 
                   decor = "Zarza De Montanchez",
                   doss = system.file("extdata", 
                                      package = "iconr"))
eds.df <- read_eds(site = "Zarza de Montanchez",
                   decor = "Zarza De Montanchez",
                   doss = system.file("extdata", 
                                      package = "iconr"))
zm <- plot_dec_grph(nodes = nds.df,
                            edges = eds.df,
                            site = "Zarza de Montanchez",
                            decor = "Zarza De Montanchez",
                            nd.var = "type",
                            lbl.size = 1.7,
                            doss = system.file("extdata", 
                                               package = "iconr"))
zm.img <- magick::image_read(zm)
out.img <- magick::image_append(c(cm1.img, zm.img))
plot(out.img)
@

@@

<<results='asis', include=FALSE, echo=FALSE, warning=FALSE>>=
library("iconr")
library("igraph")
library("magick")
@
%
The \pkg{iconr} package can be downloaded from GitHub
%
\begin{CodeChunk}
\begin{CodeInput}
R>   devtools::install_github("zoometh/iconr", build_vignettes=TRUE)
\end{CodeInput}
\end{CodeChunk}
%

@@

\subsubsection{Imgs dataframe}
The inventory of decorations is stored in the comma-separated values file \code{imgs.csv} and the tab-separated values files \code{imgs.tsv}. Any of them can be read as a dataframe (Table \ref{tab:timg}),
%
<<results='asis', echo=TRUE, warning=FALSE>>=
imgs <- read.table(system.file("extdata", "imgs.tsv", package = "iconr"),
                   sep="\t", stringsAsFactors = FALSE)
@
The field \code{imgs$idf} is the short name of the decoration, useful during statistical analysis. The primary key of each decoration is the concatenate of \code{imgs$site} and \code{imgs$decor}. These keys will allow joints with the other dataframes (nodes and edges).
<<results='asis', echo=FALSE, warning=FALSE>>=
print(xtable::xtable(imgs,
                     caption = "The studied corpus, the \\code{imgs.tsv} dataframe",
                     label = "tab:timg",
                     size=6,
                     align=c(
                       "p{0.01cm}",
                       "c",
                       ">{\\raggedright\\arraybackslash\\rule{0pt}{1em}\\addtolength{\\baselineskip}{-1.5pt}}p{2.1cm}",
                       ">{\\raggedright\\arraybackslash\\addtolength{\\baselineskip}{-1.5pt}}p{2.1cm}",
                       "p{8.5cm}")),
      table.placement="!htbp",
      include.rownames=FALSE)
@

At first the drawings dataset \citep{DiazGuardamino10} can be checked by using the information in the \code{imgs} dataframe and the \pkg{magick} package (Figure \ref{fig:dataset})
%
<<timg, echo=TRUE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.width=15, fig.asp=0.7507508, fig.cap="\\label{fig:dataset}Decoration images of the training dataset">>=
pth <- system.file("extdata", package = "iconr")
lims <- list()
for(i in 1:nrow(imgs)){
  i1 <- image_read(paste0(pth,"/",imgs$img[i]))
  lbl.txt <- paste0(paste(imgs[i,], collapse = "\n"), "\n",
                    image_info(i1)$width, "*", image_info(i1)$height, " px")
  lims[[i]] <- image_annotate(i1, lbl.txt, location = "northwest",
                              size = 28, color = "red")
}
out.img <- image_append(c(image_append(c(lims[[1]], lims[[2]], lims[[3]])),
                          image_append(c(lims[[4]], lims[[5]]))),
                        stack = TRUE)
par(mar=c(0, 0, 0, 0))
plot(out.img)
@
%
To construct a graph overlapping the decoration images listed in the \code{imgs} dataframe, the first step is to load nodes and edges dataframes.
% \begin{itemize}
% \item The \code{nodes} dataframe
% \end{itemize}

@@

<<results='asis', echo=TRUE>>=
nodes <- read.table(system.file("extdata", "nodes.csv", package = "iconr"),
                    sep=";", stringsAsFactors = FALSE)
@

@@

\subsubsection{Edges dataframe}
Analogously, the edges for the five decorations are stored in different formats in \code{edges.csv}, \code{edges.tsv}, and \code{edges.shp}. This can be read by 
<<results='asis', echo=TRUE>>=
edges <- read.table(system.file("extdata", "edges.csv", 
                                package = "iconr"),
                    sep=";",stringsAsFactors = FALSE)
@

@@

<<tedss, results='asis', echo=FALSE>>=
caption <- "Edges of Cerro Muriano 1 (from \\code{edges.csv} dataframe)"
print(xtable::xtable(edges[edges$decor == "Cerro Muriano 1", ],
                     align = "rllrrc",
                     caption=caption,
                     label="tab:teds"),
      include.rownames = FALSE,
      scalebox = 0.95,
      size = "\\renewcommand{\\arraystretch}{0.92}",
#      size = "\\small\\renewcommand{\\arraystretch}{0.989}",
      table.placement="!htbp")
@

@@

Once imgs, nodes and edges dataframes loaded, the \code{list\_dec()} function allows to create graphs for all decorations within these dataframes. The result is a list of \code{igraph} objects. As an example, the first graph of the list can be plotted with the \code{igraph::plot()} function. By default, the graphs spazialisation (\code{layout}) is based on \code{x} and \code{y} columns, when exist. This is our case since we work with geometric graphs. For a same graph, very different layouts (\emph{réalisations}) exist \citep{Mathis03}(\emp{graph drawing}) (Figure \ref{fig:igraph})

<<igraphs, echo=TRUE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.width=4, fig.height=4, fig.cap="\\label{fig:igraph}R interface. Plot of the first graph of the list (Cerro Muriano 1 decoration) with graph drawing on x, y coordinates (left), and force-directed graph drawing with nodes sized on their degree centralities (right)">>=
lgrph <- list_dec(imgs, nodes, edges)
g <- lgrph[[1]]
par(mar = c(0, 0, 0, 0), mfrow = c(1, 2))
coords <- layout.fruchterman.reingold(g)
plot(g,
     vertex.size = 14,
     vertex.label.family = "Helvetica",
     edge.color = "orange",
     vertex.label.cex = .8
)
plot(g,
     layout = coords,
     vertex.size = 4 + (degree(g)*10),
     vertex.label.family = "Helvetica",
     edge.color = "orange",
     vertex.label.cex = .8
)
@

@@

<<ibah, echo=TRUE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.width=3.5, fig.asp=1.126126, fig.cap="\\label{fig:ibah}Ibahernando decoration, node 1 overlaps node 2 and node 3">>=
par(mar=c(0,0,0,0))
nds.df <- read_nds(site = "Ibahernando", 
                   decor = "Ibahernando",
                   doss = system.file("extdata", 
                                      package = "iconr"))
eds.df <- read_eds(site = "Ibahernando",
                   decor = "Ibahernando",
                   doss = system.file("extdata", 
                                      package = "iconr"))
img.graph <- plot_dec_grph(nodes = nds.df,
                           edges = eds.df,
                           site = "Ibahernando",
                           decor = "Ibahernando",
                           lbl.size = 2,
                           doss = system.file("extdata", 
                                              package = "iconr"))
plot(magick::image_read(img.graph))
@

@@

<<ibah2, echo=TRUE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.width=8, fig.asp=0.5630631, fig.cap="\\label{fig:ibah2}Ibahernando stelae before and after the selection of node 4 (sword) graph component">>=
par(mar=c(0,0,0,0))
selected.nd <- 4
nds.df <- read_nds(site = "Ibahernando", 
                   decor = "Ibahernando",
                   doss = system.file("extdata", 
                                      package = "iconr"))
eds.df <- read_eds(site = "Ibahernando",
                   decor = "Ibahernando",
                   doss = system.file("extdata", 
                                      package = "iconr"))
l_dec_df <- contemp_nds(nds.df, eds.df, selected.nd)
Ibahernando <- plot_dec_grph(nodes = nds.df,
                             edges = eds.df,
                             site = "Ibahernando",
                             decor = "Ibahernando",
                             nd.var = "type",
                             lbl.color = "brown",
                             lbl.size=2.2,
                             doss = system.file("extdata", 
                                                package = "iconr"))
Ibahernando.img <- magick::image_read(Ibahernando)
Ibahernando.contemp <- plot_dec_grph(nodes = l_dec_df[[1]],
                                     edges = l_dec_df[[2]],
                                     site = "Ibahernando",
                                     decor = "Ibahernando",
                                     nd.var = "type",
                                     lbl.color = "brown",
                                     lbl.size=2.2,
                                     doss = system.file("extdata", 
                                                        package = "iconr"))
Ibahernando.contemp.img <- magick::image_read(Ibahernando.contemp)
out <- magick::image_append(c(Ibahernando.img, Ibahernando.contemp.img))
plot(out)
@

@@

\begin{CodeChunk}
\begin{CodeInput}
R> lgrph <- list_dec(imgs, nodes, edges)
R> named_elements(lgrph[[2]], focus = "edges", nd.var="type")
## [1] "epee-=-fibule" "bouclier-=-miroir"
## [3] "epee-=-lance" "bouclier-=-chariot_char"
## [5] "chariot_char-+-cheval" "chariot_char-+-cheval#"
## [7] "arc-=-bouclier" "arc-=-chariot_char"
## [9] "arc-=-miroir" "bouclier-=-fibule"
## [11] "bouclier-=-epee" "epee-=-miroir"
## [13] "arc-+-fleche" "chariot_char-+-roue"
## [15] "chariot_char-+-roue#"
R> as.character(named_elements(lgrph[[2]], focus = "nodes", nd.var="type"))
## [1] "chariot_char" "bouclier" "arc" "miroir"
## [5] "fibule" "epee" "lance" "cheval"
## [9] "cheval#" "roue" "roue#" "fleche"
\end{CodeInput}
\end{CodeChunk}

@@

\begin{Code}
side_plot(grp, doss, nd.var, focus = "nodes",
          nd.color = c("orange", "red"),
          nd.size = c(0.5, 1),
          ed.color = c("orange", "red"),
          ed.width = c(1, 2),
          lbl.size = 0.5)
\end{Code}

\code{side_plot()} allows to plot two decoration graphs side-by-side for elements (nodes or edges) comparisons. It take a list of two graph (\code{grp}), the path of the working directory (\code{doss}) and the name of the studied node variable (\code{nd.var}) and return an image.

@@

\begin{Code}
read_nds(site,
         decor,
         doss = getwd(),
         nodes = "nodes",
         dev = ".tsv")
\end{Code}

\code{read_nds()} is close to the \proglang{R} native \code{read.table()} function, but allows to read \emph{shapefiles} of nodes. For example, the nodes of the Torrejon Rubio 1 decoration can be read from the the \emph{shapefile}, and the 6\textsuperscript{th} node -- a sword (\code{epee}) -- displayed (Table \ref{tab:nd}).

>>>>

%% This modifies which proportion of the page must be occupied by the figure
%% or table to take the full page (with no additional text).
%% Check if the journal does not have any rule about not changing this.
\renewcommand{\floatpagefraction}{.8}
%\definecolor{shadecolor}{rgb}{1, 1, 1}
\definecolor{fgcolor}{rgb}{0, 0, 0}
\makeatletter
\def\verbatim@font{\ttfamily\slshape}
\makeatother

% \setlist[itemize]{noitemsep, topsep=0pt}
<<preliminaries, echo=FALSE, include=FALSE, cache=FALSE >>=
# options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
library("MASS")
library("extrafont") #;library("ggplot")
library("igraph")
knitr::opts_chunk$set(prompt=TRUE, fig.pos = '!htbp',
                      background = 'white', highlight=FALSE,
                      class.source = "bg-success")
options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
options(SweaveHooks=list(fig=function() par(mar=c(0, 0, 0, 0))))
# options(replace.assign=TRUE, width=90, prompt="R> ")
opts_knit$set(concordance=TRUE, self.contained=FALSE, background = "#FFFFFF")
# knitr::opts_chunk$set(fig.pos = '!htbp')
#loadfonts(device = "win")
loadfonts(device = "pdf")
@

@@

\begin{document}
%\SweaveOpts{concordance=TRUE}
%\SweaveOpts{concordance=TRUE}
%\SweaveOpts{concordance=TRUE}
% \SweaveOpts{concordance=TRUE}
% \Sweave2knitr("article_rvTH14_1.Rnw")
% \SweaveOpts{concordance=TRUE}

@@

Symbolic practices is a characteristic trait of human societies. Even discussed, such practices seems to start between 233,000 to 800,00 BC \citep{dErrico00}, covering more than 97\% of total human societies time span. Symbolism cover a large range of practices, from ochre deposit in a tomb, to menhir alignements, passing through wall fresco. This latter, what might be called "iconographical practices", probably shows the most complex and interesting testimonies of past societies symbolism. For decades, its study was linked to history of religion because commonly seen as closely linked to cultual practices and believes. Since the \textit{New Archaeology} developpement during the 60's \citep{Clarke14}, symbolic expressions start to be studied with the same formal methods (statistics, seriations, distribution maps, etc.) as any another aspect of social organisation: settlement patterns, tools \emph{chaine op\'eratoire}, susbsitence strategies, etc. \citep{Renfrew91}. But unlike many aspects of the material culture where technological requirements and technical efficiency determine the choice of the raw material and the shape of the object -- a flint blade for cutting, a pottery for containing, a house for living --, the function of an iconographic composition cannot be drawn directly from itself. Furthermore, there is no specific and controled vocabularies to describe hand-made images. There is always a difficulty to decribe an iconographical content with alphanumerical signs. Whether these last decades study of ancient iconography had  undergone significative improvements at the site scale -- with GIS/database statistics, paleoenvironmental restitutions, etc. -- and at the sign scale with the development of archaeological sciences -- radiocarbon dating, use-wear analysis, elemental analysis, etc. --, these improvement do not necessarly help to understand the semantic content of the iconography.\\
Semantics or semiotics can be defined as a system of conventional features -- called signs -- organised also in conventional manners. Until our days, formal methods to study ancient iconography semantics have been mostlty been grounded  -- explicitly or not -- on the prime principle of Saussurian linguistic: the '\emph{linearity of the signifier}' \citep{Saussure89}.
Writing is one of the most rational semiographical system with a clear distinction between signified and signifier and the development of the signified on a linear axis. Even if we do not understand the meaning of the signs, writing can easily be modelled with Graph theory and recurent patterns can be identified. For example, the 3-letters word "\code{art}" can be modelled with three vertices (\code{a}, \code{r}, \code{t}) and two edges (one between \code{a} and \code{r}, the other between \code{r} and \code{t}). In \proglang{R}, these features, concatenated in this order with a \code{paste0()}, is \code{art}, and not \code{rat} (Figure \ref{fig:a_r_t}).

<<a_r_t, echo=FALSE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.width=1.25, fig.height=0.25, fig.cap="Concatenate of 'a', 'r' and 't' graphical units is 'art'">>=
par(mar=c(0.1,0.1,0.1,0.1) )
g <- graph_from_literal(a-+r, r-+t)
plot(g,
     layout = cbind(seq(-1,1,length.out = gorder(g)), 0),
     vertex.color="white",
     vertex.frame.color="white",
     vertex.size=40,
     vertex.label.cex=1,
     vertex.label.color="black",
     vertex.label.family="Courier",
     edge.arrow.size=0.5,
     xlim=c(-0.94,0.94),
     ylim=c(-1,-1))
@

@@

<<fratar, echo=FALSE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.width=1.4, fig.height=1.4, fig.cap="Potential spatial relations between graphical units">>=
par(mar=c(0,0,0,0))
ugs <- data.frame(name=c("a","r","t"))
lks <- data.frame(from=c("a", "r", "t"), to=c("r", "t", "a"), type=c("=","=","="))
g <- igraph::graph_from_data_frame(lks, directed=FALSE, vertices= ugs)
plot(g,
     vertex.color="white",
     vertex.frame.color="white",
     vertex.size=80,
     vertex.label.cex=1,
     vertex.label.color="black",
     vertex.label.family="Courier",
     xlim=c(-0.65,0.65),
     ylim=c(-0.65,0.65)
)
@

Because of the inherent variability of iconography, and because graphical and spatial proximities between graphical units are generally not quantified, applying the Saussurian model to any structured prehistorical graphical content had led to considerable problems:
\begin{itemize}
\item unexplicit spatial groupings of graphical units -- like graphical units grouped into \emph{figures},  \emph{figures} grouped into \emph{patterns}, \emph{patterns} grouped into \emph{motives}, etc. -- with tedious number of groups
\item consistency, proximities and relationships between these groups are often implicit and not quantified
\item studies develop proper descriptive vocabularies, singular relationships of grouping, ido\-syn\-crat\-ic methods at site-dependent or period-dependent scales
\end{itemize}

@@

Whether this theoretical process includes image binarization, polygonization and \emph{birel} analysis, a simplier solution will be to create directly centroids on the GUs and to draw the edges manually. Then \emph{x} and \emph{y} coordinates of nodes can be easily retrieved.

@@

<<dec.1.4, echo=FALSE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.width=8, fig.asp=0.5, fig.cap="Cerro Muriano 1 and Zarza de Montanchez decorations.">>=
library(iconr)
imgs <- read.table(system.file("extdata", "imgs.tsv", package = "iconr"),
                   sep="\t", stringsAsFactors = FALSE)
sites = c("Cerro Muriano", "Zarza de Montanchez")
decors = c("Cerro Muriano 1", "Zarza De Montanchez")
par(mfrow = c(1, 2))
for(i in 1:2) {
  nds.df <- read_nds(sites[i], decors[i],
                     dir = system.file("extdata", package = "iconr"))
  eds.df <- read_eds(sites[i], decors[i],
                     dir = system.file("extdata", package = "iconr"))
  plot_dec_grph(nds.df, eds.df, imgs,
                sites[i], decors[i],
                nd.var = "type",
                dir = system.file("extdata", package = "iconr"),
                img.format = NULL)
}
@

@@

<<include=FALSE, echo=FALSE, warning=FALSE>>=
library("iconr")
library("igraph")
library("magick")
@
%
The \pkg{iconr} package can be downloaded from GitHub
%
\begin{CodeChunk}
\begin{CodeInput}
R> devtools::install_github("zoometh/iconr", build_vignettes=TRUE)
\end{CodeInput}
\end{CodeChunk}
%

@@

\subsubsection{Imgs dataframe}
The inventory of decorations is stored in the comma-separated values file \code{imgs.csv} and the tab-separated values files \code{imgs.tsv}. Any of them can be read as a dataframe (Table \ref{tab:timg}),
%
<<echo=TRUE, warning=FALSE>>=
imgs <- read.table(system.file("extdata", "imgs.tsv", package = "iconr"),
                   sep="\t", stringsAsFactors = FALSE)
@
The field \code{imgs$idf} is the short name of the decoration, useful during statistical analysis. The primary key of each decoration is the concatenate of \code{imgs$site} and \code{imgs$decor}. These keys will allow joints with the other dataframes (nodes and edges).
<<results='asis', echo=FALSE, warning=FALSE>>=
print(xtable::xtable(imgs,
                     caption = "The studied corpus, the \\code{imgs.tsv} dataframe",
                     label = "tab:timg",
                     size=6,
                     align=c(
                       "p{0.01cm}",
                       "c",
                       ">{\\raggedright\\arraybackslash\\rule{0pt}{1em}\\addtolength{\\baselineskip}{-1.5pt}}p{2.1cm}",
                       ">{\\raggedright\\arraybackslash\\addtolength{\\baselineskip}{-1.5pt}}p{2.1cm}",
                       "p{8.5cm}")),
      table.placement="!htbp",
      include.rownames=FALSE)
@

At first the drawings dataset \citep{DiazGuardamino10} can be checked by using the information in the \code{imgs} dataframe and the \pkg{magick} package (Figure \ref{fig:dataset})
%
<<dataset, echo=TRUE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.width=15, fig.asp=0.7507508, fig.cap="Decoration images of the training dataset">>=
ims <- list()
for(i in 1:nrow(imgs)){
  im <- image_read(system.file("extdata", imgs$img[i], package = "iconr"))
  lbl.txt <- paste0(paste(imgs[i,], collapse = "\n"), "\n",
                    image_info(im)$width, "*", image_info(im)$height,
                    " px")
  ims[[i]] <- image_annotate(im, lbl.txt, location = "northwest",
                             size = 28, color = "red")
}
out.img <- image_append(c(image_append(c(ims[[1]], ims[[2]], ims[[3]])),
                          image_append(c(ims[[4]], ims[[5]]))),
                        stack = TRUE)
par(mar=c(0, 0, 0, 0))
plot(out.img)
@
%
To construct a graph overlapping the decoration images listed in the \code{imgs} dataframe, the first step is to load nodes and edges dataframes.
% \begin{itemize}
% \item The \code{nodes} dataframe
% \end{itemize}

@@

<<echo=TRUE>>=
nodes <- read.table(system.file("extdata", "nodes.csv",
                                package = "iconr"),
                    sep=";", stringsAsFactors = FALSE)
@

@@

\subsubsection{Edges dataframe}
Analogously, the edges for the five decorations are stored in different formats in \code{edges.csv}, \code{edges.tsv}, and \code{edges.shp}. This can be read by
<<echo=TRUE>>=
edges <- read.table(system.file("extdata", "edges.csv",
                                package = "iconr"),
                    sep=";",stringsAsFactors = FALSE)
@

@@

<<tedss, results='asis', echo=FALSE>>=
caption <- "Edges of Cerro Muriano 1 (from \\code{edges.csv} dataframe)"
print(xtable::xtable(edges[edges$decor == "Cerro Muriano 1", ],
                     align = "rllrrc",
                     caption=caption,
                     label="tab:teds"),
      include.rownames = FALSE,
#      scalebox = 0.95,
#      size = "\\renewcommand{\\arraystretch}{0.92}",
#      size = "\\small\\renewcommand{\\arraystretch}{0.989}",
      table.placement="!htbp")
@

@@

Once imgs, nodes and edges dataframes loaded, the \code{list\_dec()} function allows to create graphs for all decorations within these dataframes. The result is a list of \code{igraph} objects. As an example, the first graph of the list can be plotted with the \code{igraph::plot()} function. By default, the graphs spazialisation (\code{layout}) is based on \code{x} and \code{y} columns, when exist. This is our case since we work with geometric graphs. For a same graph, very different layouts (\emph{réalisations}) exist \citep{Mathis03}(\emp{graph drawing}) (Figure \ref{fig:igraphs})

<<igraphs, echo=TRUE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.asp=0.5, fig.height=4, fig.cap="R interface. Plot of the first graph of the list (Cerro Muriano 1 decoration) with graph drawing on x, y coordinates (left), and force-directed graph drawing with nodes sized on their degree centralities (right)">>=
lgrph <- list_dec(imgs, nodes, edges)
g <- lgrph[[1]]
par(mar = c(0, 0, 0, 0), mfrow = c(1, 2))
plot(g,
     vertex.size = 14,
     vertex.label.family = "Helvetica",
     edge.color = "orange",
     vertex.label.cex = .8
)
plot(g,
     layout = layout.fruchterman.reingold(g),
     vertex.size = 4 + (degree(g)*10),
     vertex.label.family = "Helvetica",
     edge.color = "orange",
     vertex.label.cex = .8
)
@

@@

<<ibah, echo=TRUE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.width=3.2, fig.asp=1.126126, fig.cap="Ibahernando decoration{:} node 1 overlaps node 2 and node 3.">>=
site = "Ibahernando"
decor = "Ibahernando"
dataDir = system.file("extdata", package = "iconr")
nds.df <- read_nds(site, decor, dataDir)
eds.df <- read_eds(site, decor, dataDir)
par(mar=c(0,0,0,0))
plot_dec_grph(nds.df, eds.df, imgs,
              site, decor, dir = dataDir,
              lbl.size = 0.7, img.format = NULL)
@

@@

%<<ibah2, echo=TRUE, message=FALSE, warning=FALSE>>=
%nds.df <- read_nds(site = "Ibahernando",
%                   decor = "Ibahernando",
%                   dir = system.file("extdata", package = "iconr"))
%eds.df <- read_eds(site = "Ibahernando",
%                   decor = "Ibahernando",
%                   dir = system.file("extdata", package = "iconr"))
%l_dec_df <- contemp_nds(nds.df, eds.df, selected.nd = 4)
%l.nodes = list(nds.df, l_dec_df[[1]])
%l.edges = list(eds.df, l_dec_df[[2]])
%out.file.name = paste0("figure/ibah2_", c("a", "b"), ".pdf")
%img.files <- character(0)
%for(i in 1:2) {
%  img.files[i] <- plot_dec_grph(nodes = l.nodes[[i]],
%                                  edges = l.edges[[i]],
%                                  site = "Ibahernando",
%                                  decor = "Ibahernando",
%                                  nd.var = "type",
%                                  lbl.color = "brown",
%                                  lbl.size = 0.7,
%                                  dir = system.file("extdata",
%                                                     package = "iconr"),
%#                                  img.format = "pdf")
%                                  out.file.name = out.file.name[i])
%}
%@
%\begin{figure}[!htbp]
%  \centering
%  \centering
%  \includegraphics[width=0.48\textwidth]{\Sexpr{img.files[1]}}\quad
%  \includegraphics[width=0.48\textwidth]{\Sexpr{img.files[2]}}
%  \caption{Ibahernando stelae before and after the selection of node 4 (sword) graph component.}
%  \label{fig:ibah2}
%\end{figure}

<<ibah2, echo=TRUE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.width=8, fig.asp=0.5, fig.cap="Ibahernando stelae before and after the selection of node 4 (sword) graph component">>=
site = "Ibahernando"
decor = "Ibahernando"
dataDir = system.file("extdata", package = "iconr")
nds.df <- read_nds(site, decor, dataDir)
eds.df <- read_eds(site, decor, dataDir)
l_dec_df <- contemp_nds(nds.df, eds.df, selected.nd = 4)
par(mfrow = c(1, 2), mar = c(0, 0, 0, 0))
plot_dec_grph(nds.df, eds.df, imgs,
              site, decor, dir = dataDir,
              nd.var = "type", lbl.color = "brown", lbl.size = 0.7,
              img.format = NULL)
plot_dec_grph(l_dec_df$nodes, l_dec_df$edges, imgs,
              site, decor, dir = dataDir,
              nd.var = "type", lbl.color = "brown", lbl.size = 0.7,
              img.format = NULL)
@

@@

<<named_elements, echo=TRUE>>=
lgrph <- list_dec(imgs, nodes, edges)
named_elements(lgrph[[2]], focus = "edges", nd.var="type")
named_elements(lgrph[[2]], focus = "nodes", nd.var="type")
@

@@

\begin{Code}
side_plot(grp, dir, nd.var, focus = "nodes",
          nd.color = c("orange", "red"),
          nd.size = c(0.5, 1),
          ed.color = c("orange", "red"),
          ed.width = c(1, 2),
          lbl.size = 0.5)
\end{Code}

\code{side_plot()} allows to plot two decoration graphs side-by-side for elements (nodes or edges) comparisons. It take a list of two graph (\code{grp}), the path of the working directory (\code{dir}) and the name of the studied node variable (\code{nd.var}) and return an image.

@@

\begin{Code}
read_nds(site,
         decor,
         dir = getwd(),
         nodes = "nodes",
         dev = ".tsv")
\end{Code}

\code{read_nds()} is close to the \proglang{R} native \code{read.table()} function, but allows to read \emph{shapefiles} of nodes. For example, the nodes of the Torrejon Rubio 1 decoration can be read from the \emph{shapefile} (Table \ref{tab:nd} displays two nodes as example).

<<nd0, echo=TRUE>>=
nds.df <- read_nds(site = "Torrejon Rubio",
                   decor = "Torrejon Rubio 1",
                   dir = system.file("extdata", package = "iconr"),
                   dev = ".shp")
@
<<nd, results='asis', echo=FALSE>>=
print(xtable::xtable(nds.df[5:6, ],
                     caption = "Torrejon Rubio 1 decoration nodes 5 (fibula) and 6 (sword).",
                     label = "tab:nd",
                     size = 7),
      include.rownames = FALSE,
      table.placement = "!htbp")
@
